From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-15 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
very minor changes; pushing just so I can pull new changes!

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the scalar value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more matter at this point. Likewise, if the scalar value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, we'll refer to this as the property of \textit{point-wise comparison}. Point-wise comparison requires a coordinate-invariant density. As the derivation in Section~\ref{deterministic} shows, to guarantee the existence of an invariant density, we must define state space over a symplectic manifold. In other words, to generically have a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold.

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Consequently, within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues for the fundamental physical significance of the phase space density. Since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics. The syntactic symmetry between position and momentum in the mathematical formalism does not entail that these quantities are equally physically significant. 
%\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.}
% Canonical coordinates are physically significant because they allow us to define an invariant phase space density.
% % This latter claim might seem puzzling, since insofar as we need to use canonical coordinates to define the density, this seems to speak in favor of the canonical coordinates being more fundamental.

@@

Section~\ref{fundamentality} characterized fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports point-wise density comparisons, we have encountered two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See~\ref{gauge} for discussion.} Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state, in virtue of representing the distribution.

Section~\ref{infinitesimal} demonstrated a third reason for viewing the density as fundamental: by defining the density, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another account of why this mathematical structure arises for classical systems. Collectively, these considerations speak in favor of viewing the invariant phase space density as not only physically significant but also fundamental. 
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

Yet, as Section~\ref{dependence} explained, the density is always coordinate-dependent, even when coordinate-invariant. In other words, there is no coordinate-independent (or coordinate-free definition) of the invariant phase space density. Affording the density fundamental significance therefore places us at odds with those who take coordinate-independence as a necessary condition for fundamentality. Proponents of this necessary condition might respond by noting that although the density is not coordinate-independent, the distribution is. For the distribution is equivalent to the classical state, which is a thing in the world, i.e. an ontic entity. We grant that the distribution is ontically fundamental in a sense that does not accrue to the density (which represents the distribution). However, we maintain that the density itself is of fundamental significance, both epistemically and physically. For it is only through the density that we gain epistemic access to the distribution. Although any particular choice of coordinates is arbitrary, we need to make some choice in order to quantitatively measure the distribution through experiments. Additionally, defining the density is necessary for deriving symplectic structure, as shown in Section~\ref{infinitesimal}. Insofar as our opponent views the symplectic structure of Hamiltonian mechanics as fundamental, they must view the density as equally fundamental---despite its coordinate-dependence. Likewise, insofar as our opponent refuses to grant the density fundamental status, they will have to withhold this same status from the symplectic structure. Hence, the coordinate-independent notion of fundamentality needs to be supplemented with a coordinate-dependent, epistemic notion of fundamentality. This latter notion is no less important, and it shows that it is a mistake to desire a completely ``intrinsic" formulation of a physical theory.
% % Further claim: the ontic, coordinate-independent notion of fundamentality is insufficient for doing physics. We need a corresponding epistemic notion of fundamentality, since we units, state variables, and the density in order to connect the distribution to experiment. One can argue that any particular choice of state variables or units or coordinates for the density is arbitrary, but we need some arbitrary choice in order to quantitatively measure the distribution through experiments.
% % Someone might also worry that the density cannot be more fundamental than the state variables, since we define the density over state variables. In response, we can say the following: whereas the density is coordinate-invariant, the state variables are not. So the state variables are more arbitrary than the density. Note that we do not directly measure states of particles. We measure quantities. So the only way to measure the distribution is to express it as a density in terms of some quantities, i.e. in terms of some state variables. So we need the state variables to give epistemic access to the distribution, via the density. So the state variables themselves are not important (e.g. we can choose the unit basically however we want, and the conjugate momentum has even more freedom, due to gauge redundancy). The physically important thing is access to the distribution, which we gain through the density. The classical system is made up of pieces and the density lets us say how many pieces we have in various states.
% A defender of the necessity of coordinate-independence for fundamentality might respond as follows: 
% % [To do: provide literature references for this and motivating quotations; likewise for earlier remark about intrinsicality: Field, North, Chen]

>>>>

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more matter at this point. Likewise, if the value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, we'll refer to this as the property of \textit{point-wise comparison}. Point-wise comparison requires a coordinate-invariant density. As the derivation in Section~\ref{infinitesimal} shows, to guarantee the existence of an invariant density, we must define state space over a symplectic manifold. In other words, to generically have a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold.

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Consequently, within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues for the fundamental physical significance of the phase space density. Since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than the choice of state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics. The syntactic symmetry between position and momentum in the mathematical formalism does not entail that these quantities are equally physically significant. 
%\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.}
% Canonical coordinates are physically significant because they allow us to define an invariant phase space density.
% % This latter claim might seem puzzling, since insofar as we need to use canonical coordinates to define the density, this seems to speak in favor of the canonical coordinates being more fundamental.

@@

Section~\ref{fundamentality} characterized fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports point-wise density comparisons, we have encountered two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state, in virtue of representing the distribution.

Section~\ref{infinitesimal} demonstrated a third reason for viewing the density as fundamental: by defining the density, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another account of why this mathematical structure arises for classical systems. Collectively, these considerations support interpreting the invariant phase space density as not only physically significant but also fundamental. 
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

Yet, as Section~\ref{dependence} explained, the density is always coordinate-dependent, even when coordinate-invariant. In other words, there is no coordinate-independent (or coordinate-free definition) of the invariant phase space density. Affording the density fundamental significance therefore places us at odds with those who take coordinate-independence as a necessary condition for fundamentality.

Proponents of this necessary condition might respond by noting that although the density is not coordinate-independent, the distribution is. For the distribution is equivalent to the classical state, which is a thing in the world, i.e. an ontic entity. We grant that the distribution is ontically fundamental in a sense that does not accrue to the density (which represents the distribution). However, we maintain that the density itself is of fundamental significance, both epistemically and physically. For it is only through the density that we gain epistemic access to the distribution. Although the choice of coordinates is fairly arbitrary, we need to make some choice in order to quantitatively measure the distribution through experiments. Additionally, defining the density is necessary for deriving symplectic structure, as shown in Section~\ref{infinitesimal}. Insofar as our opponent views the symplectic structure of Hamiltonian mechanics as fundamental, they must view the density as equally fundamental---despite its coordinate-dependence. For the same reason, insofar as our opponent refuses to grant the density fundamental status, they will have to withhold this same status from the symplectic structure. Hence, the coordinate-independent notion of fundamentality needs to be supplemented with a coordinate-dependent, epistemic notion of fundamentality. This latter notion is no less important, and it shows that it is a mistake to desire a completely ``intrinsic" formulation of a physical theory.
% % Further claim: the ontic, coordinate-independent notion of fundamentality is insufficient for doing physics. We need a corresponding epistemic notion of fundamentality, since we units, state variables, and the density in order to connect the distribution to experiment. One can argue that any particular choice of state variables or units or coordinates for the density is arbitrary, but we need some arbitrary choice in order to quantitatively measure the distribution through experiments.
% % Someone might also worry that the density cannot be more fundamental than the state variables, since we define the density over state variables. In response, we can say the following: whereas the density is coordinate-invariant, the state variables are not. So the state variables are more arbitrary than the density. Note that we do not directly measure states of particles. We measure quantities. So the only way to measure the distribution is to express it as a density in terms of some quantities, i.e. in terms of some state variables. So we need the state variables to give epistemic access to the distribution, via the density. So the state variables themselves are not important (e.g. we can choose the unit basically however we want, and the conjugate momentum has even more freedom, due to gauge redundancy). The physically important thing is access to the distribution, which we gain through the density. The classical system is made up of pieces and the density lets us say how many pieces we have in various states.
% A defender of the necessity of coordinate-independence for fundamentality might respond as follows: 
% % [To do: provide literature references for this and motivating quotations; likewise for earlier remark about intrinsicality: Field, North, Chen]
